Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2023-12-21 doc/manuscript/translocation_bioarxiv_SI.tex
Compile PDF versions of manuscript

Swab extracts were analyzed using standard Bd DNA extraction and qPCR
methods (74), and extracts were analyzed singly instead of in triplicate
(87). For analysis of swabs collected during 2005--2014, we used
standards developed from known concentrations of zoospores (74), and
after 2014, we used standards based on single ITS1 PCR amplicons (88).
Based on paired comparisons between samples analyzed using both types of
standards, Bd in the study area has an average of 60 ITS1 copies per
zoospore. To express all qPCR results as the number of ITS1 copies,
starting quantities obtained using the zoospore standard (measured as
``zoospore equivalents'') were multiplied by 60. In addition, all qPCR
quantities (regardless of standard) were multiplied by 80 to account for
the fact that DNA extracts from swabs were diluted 80-fold during
extraction and PCR (33).

@@

We estimated survival and recruitment for each site using open
population CMR models based on (41). For each individual \(i=1, ..., M\)
on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
histories of \(M\) individuals are modeled, although only \(N_s\)
individuals were captured. This parameter expanded data augmentation
allows us to capture the possibility that undetected individuals may
have recruited into the adult population (89). Here, \(M\) was chosen to
be three times the number of observed individuals (\(3N_s\)) to be
considerably greater than our prior guess of \(N_s\).

@@

We computed the likelihood of each individual capture history using the
forward algorithm, and we estimated the latent states using the
forward-backward algorithm (41, 90).

All of the code to specify and fit the model in Stan is available in the
open source mrmr package (67).

@@

These models were fit using the \texttt{stan\_glmer()} function in the
rstanarm package, with default priors described below (71). These priors
are vague, but include data-dependent scaling as follows to account for
different input variable scales. However, because we standardized all
predictor variables similarly to have equal variance (by centering and
dividing by twice the sample standard deviation), the resulting priors
are identical. Specifically, we have:

@@

The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
rstanarm is a zero-mean Gaussian, where the covariance matrix is
constructed from a correlation matrix with an LKJ prior, and a vector of
variance parameters -- the decomposition of variance prior with unit
regularization, concentration, shape, and scale parameters (91).

We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
with four parallel chains, each run for 10,000 iterations, discarding
the first half of each chain as warm-up draws (71). We used Rhat
statistics and trace plots to verify convergence. We considered models
with different subsets of fixed and random effects, and used approximate
leave-one-out cross validation to identify the best model (72).

@@

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
Using Approximate Bayesian Computing and rejection sampling (92), we
identified the top 2\% of trajectories (i.e., 1000 trajectories) that
minimized the sum of squared errors between the observed and predicted
data. The yearly \(\sigma_{J_1}\) values associated with these ``best''
trajectories represented an approximate posterior distribution (93).
Using these best fit trajectories, we assessed whether our model could
qualitatively describe the patterns of recovery in the observed data for
population 70550.

@@

Our study design, in which we compared frog genomes in naive and
recovering populations, required sampling populations across the range
of MYL frogs in the southern Sierra Nevada. This is due to the fact that
very few MYL frog populations remain in the naive state, and those that
do are scattered across a wide latitudinal range, from Yosemite National
Park in the north to southern Kings Canyon National Park in south. To
minimize potential confounding effects caused by known variation in frog
genotypes across latitude (49), we selected sampling sites such that
both population types were represented across similar latitudinal ranges
(Fig. 1).

@@

In \textbf{Results-Frog evolution in response to Bd}, we describe the
stringent set of outlier variants (identified using a
Bonferroni-corrected p-value of 0.01). A liberal set of outlier
variants, identified using a Bonferroni-corrected p-value of 0.05,
included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
across 16 contigs. We used this liberal set to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. To do this, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. We then conducted a
statistical overrepresentation test (Fisher's exact test) using Blast2GO
(94) to compare the 30 unique outlier genes to the complete set of genes
in our target transcriptome. We repeated this process for the set of 33
genes located in the \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows.

@@

To characterize general patterns of genetic diversity between naive and
persistent populations, we calculated individual heterozygosity using
VCFtools (85), and compared average heterozygosity using a Wilcoxon
rank-sum test. Additionally, we used VCFtools to calculate
per-population \(\pi\) in 100kb sliding windows along the genome, and in
each of the 10 outlier windows identified in the splined window
analysis. We compared per-population average \(\pi\) values using
Wilcoxon rank-sum tests.

Average (i) individual-level heterozygosity and (ii) population-level
nucleotide diversity (\(\pi\)) were similar between naive and recovering
populations (Figure~\ref{fig-violinplot-heterozy};
Figure~\ref{fig-boxplot-pi}). The analysis of \(\pi\) across the 10
outlier windows indicates that within each of the windows there is
considerable variation in \(\pi\) between populations and between
population clusters (Figure~\ref{fig-barchart-pi-by-windowsite}).

@@

\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
outlier region containing MHC Class I Classical and MHC Class III gene
regions. The plot was created with SimpleSynteny (95) using
\emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank accession
GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices indicate
the location of SNP outliers in the TCF19 and VARS1 genes. Gap sizes for
each contig representation are labeled.}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/het_by_status_violin_plot.png}

@@

\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
heterozygosity for the Bd-naive and recovering populations included in
the frog evolution study. Individual data points are represented by
their corresponding site number (from
Fig. 1). A Wilcoxon signed-rank test showed
no significant difference between the two groups.}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_site.png}

@@

\caption{\label{fig-boxplot-pi}For each study population in the frog
evolution study (from Fig. 1), boxplots
showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
along the genome. The y-axis has been log\textsubscript{10}-transformed
for display purposes. Sites are color-coded as ``Bd-naive'' or
``recovering'', and vertical dashed lines show the genetic breaks
between frog populations as described in
Figure~\ref{fig-selectionresults} A). Within the genetic clusters,
mean \(\pi\) is significantly different between sites (p \textless0.001
Wilcoxon rank-sum test).}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/perwindow_persite_pi.png}

@@

\caption{\label{fig-barchart-pi-by-windowsite}For each study population
in the frog evolution study, average \(\pi\) values by population
calculated for each of the 10 outlier windows identified in the splined
window analysis (A-J). Sites are color-coded as ``Bd-naive'' or
``recovering'', and vertical dashed lines show the genetic breaks
between frog populations as described in
Figure~\ref{fig-selectionresults} A. Gene annotations for the outlier
regions are as follows: (A) C6, C7, (B) DDX10, ZBTB24, (C) SULTR-like,
TRAF3IP2, VGLL3, EXOC1, (D) FLOT1, TUBB2B, TUBB, MDC1, CCHCR1,
TCF19\emph{, TRIM29, HSP70, LSM2, VARS,} (E) GCC2, CFAP251, PEG10, (F)
ERO1A, GVINP1, (G) ERO1A, GVINP1, (H) PPP1R12A, TSPAN4, PAWR, MFRP, MAX,
PPP6R3, (I) C6H5ORF22, PKS6, BSPRY, MPV17, and (J) CAD, ATRAID, GPN1.
Note that windows F and G are in the same gene region. * by gene names
indicate that an outlier SNP is also present in this gene . (K) Average
\(\pi\) values of naive and recovering groups for each of the outlier
windows shown in A-J. }

>>>>

Swab extracts were analyzed using standard Bd DNA extraction and qPCR
methods \citep{boyle2004}, and extracts were analyzed singly instead of
in triplicate \citep{kriger2006}. For analysis of swabs collected during
2005--2014, we used standards developed from known concentrations of
zoospores \citep{boyle2004}, and after 2014, we used standards based on
single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
between samples analyzed using both types of standards, Bd in the study
area has an average of 60 ITS1 copies per zoospore. To express all qPCR
results as the number of ITS1 copies, starting quantities obtained using
the zoospore standard (measured as ``zoospore equivalents'') were
multiplied by 60. In addition, all qPCR quantities (regardless of
standard) were multiplied by 80 to account for the fact that DNA
extracts from swabs were diluted 80-fold during extraction and PCR
\citep{vredenburg2010}.

@@

We estimated survival and recruitment for each site using open
population CMR models based on \citep{joseph2018}. For each individual
\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
the individual was not detected, and \(o_{i, j}=2\) if the individual
was detected. Capture histories of \(M\) individuals are modeled,
although only \(N_s\) individuals were captured. This parameter expanded
data augmentation allows us to capture the possibility that undetected
individuals may have recruited into the adult population
\citep{royle2012}. Here, \(M\) was chosen to be three times the number
of observed individuals (\(3N_s\)) to be considerably greater than our
prior guess of \(N_s\).

@@

We computed the likelihood of each individual capture history using the
forward algorithm, and we estimated the latent states using the
forward-backward algorithm \citep{zucchini2009, joseph2018}.

All of the code to specify and fit the model in Stan is available in the
open source mrmr package \citep{joseph2019}.

@@

These models were fit using the \texttt{stan\_glmer()} function in the
rstanarm package, with default priors described below
\citep{rstanarm2022}. These priors are vague, but include data-dependent
scaling as follows to account for different input variable scales.
However, because we standardized all predictor variables similarly to
have equal variance (by centering and dividing by twice the sample
standard deviation), the resulting priors are identical. Specifically,
we have:

@@

The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
rstanarm is a zero-mean Gaussian, where the covariance matrix is
constructed from a correlation matrix with an LKJ prior, and a vector of
variance parameters -- the decomposition of variance prior with unit
regularization, concentration, shape, and scale parameters
\citep{lewandowski2009}.

We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
with four parallel chains, each run for 10,000 iterations, discarding
the first half of each chain as warm-up draws \citep{rstanarm2022}. We
used Rhat statistics and trace plots to verify convergence. We
considered models with different subsets of fixed and random effects,
and used approximate leave-one-out cross validation to identify the best
model \citep{vehtari2016}.

@@

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
Using Approximate Bayesian Computing and rejection sampling
\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
1000 trajectories) that minimized the sum of squared errors between the
observed and predicted data. The yearly \(\sigma_{J_1}\) values
associated with these ``best'' trajectories represented an approximate
posterior distribution \citep{beaumont2010}. Using these best fit
trajectories, we assessed whether our model could qualitatively describe
the patterns of recovery in the observed data for population 70550.

@@

Our study design, in which we compared frog genomes in naive and
recovering populations, required sampling populations across the range
of MYL frogs in the southern Sierra Nevada. This is due to the fact that
very few MYL frog populations remain in the naive state, and those that
do are scattered across a wide latitudinal range, from Yosemite National
Park in the north to southern Kings Canyon National Park in south. To
minimize potential confounding effects caused by known variation in frog
genotypes across latitude \citep{byrne2023}, we selected sampling sites
such that both population types were represented across similar
latitudinal ranges (Fig. 1).

@@

In \textbf{Results-Frog evolution in response to Bd}, we describe the
stringent set of outlier variants (identified using a
Bonferroni-corrected p-value of 0.01). A liberal set of outlier
variants, identified using a Bonferroni-corrected p-value of 0.05,
included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
across 16 contigs. We used this liberal set to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. To do this, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. We then conducted a
statistical overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005} to compare the 30 unique outlier genes to the
complete set of genes in our target transcriptome. We repeated this
process for the set of 33 genes located in the
\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.

@@

To characterize general patterns of genetic diversity between naive and
recovering populations, we conducted three analyses. First, we
calculated heterozygosity for each sampled frog using VCFtools
\citep{danecek2011}. Second, to characterize genome-wide patterns of
diversity, we used VCFtools to calculate nucleotide diversity (\(\pi\))
in 100kb sliding windows along the genome for each population. Third, we
calculated average \(\pi\) per population within each of the 9 outlier
windows identified in the splined window analysis.

Average individual-level heterozygosity and genome-wide population-level
\(\pi\) were similar between the naive and recovering groups
(Figure~\ref{fig-violinplot-heterozy};
Figure~\ref{fig-boxplot-genomewide-pi-by-pop}). Within each of the 9
outlier windows, average \(\pi\) shows considerable variation between
populations (Figure~\ref{fig-boxplot-pi-by-windowpop}) and no obvious
patterns between the naive and recovering groups.

@@

\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
outlier region containing MHC Class I Classical and MHC Class III gene
regions. The plot was created with SimpleSynteny \citep{veltri2016}
using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
sizes for each contig representation are labeled.}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/violin_plot_heterozy_by_group.png}

@@

\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
heterozygosity for the Bd-naive and recovering populations included in
the frog evolution study. Individual data points are represented by
their corresponding site number (from
Fig. 1).}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_genomewide_pi_by_pop.png}

@@

\caption{\label{fig-boxplot-genomewide-pi-by-pop}For each study
population in the frog evolution study (from
Fig. 1), boxplots showing nucleotide
diversity (\(\pi\)) within 100kb sliding windows along the genome. The
y-axis has been log\textsubscript{10}-transformed for display purposes.
Sites are color-coded as ``Bd-naive'' or ``recovering'', and vertical
dashed lines show the genetic breaks between frog populations as
described in Figure~\ref{fig-selectionresults} A). Within the genetic
clusters, mean \(\pi\) is significantly different between sites (p
\textless0.001 Wilcoxon rank-sum test).}

@@

{\centering \includegraphics[width=0.85\textwidth]{figures/boxplot_pi_by_windowpop.png}

@@

\caption{\label{fig-boxplot-pi-by-windowpop}For each of the 9 outlier
windows identified by the overlapping splined windows for FST and
\$\textbackslash pi\$, boxplots showing average \$\textbackslash pi\$
for naive and recovering populations. The corresponding gene annotations
for each window are as follows: (1) C6, C7, (2) DDX10, ZBTB24, (3)
SULTR-like, TRAF3IP2, VGLL3, EXOC1, (4) FLOT1, TUBB, MDC1, CCHCR1,
TCF19, HSP70, LSM2, VARS1, (5) GCC2, CFAP251, PEG10, (6) ERO1A, GVINP1,
(7) PPP1R12A, TSPAN4, PAWR, MFRP, MAX, PPP6R3, (8) C6H5ORF22, PKS6,
BSPRY, MPV17, and (9) CAD, ATRAID, GPN1.}
